---
title: "Modified Macromolecular Rate Theory"
output: html_document
date: "26 March, 2023"
author: "Charlotte J. Alster, ..."
editor_options: 
  chunk_output_type: console
---

The following code fits the modified Macromolecular Rate Theory on data for log transformed growth rate over temperature. The input file is outputted by the model-fitting.Rmd script.
```{r}
library(tidyr)
library(plyr)
library(dplyr)
library(broom)
library(ggplot2)
```

Reading in the data

```{r}
setwd(".../.../...")
temp_growth <- read.csv(".../.../...") # this file is read out from model-fitting.Rmd

# removing -Inf values
temp_growth <- filter(temp_growth, ln_r != -Inf)
```

Creating subsetted data frames for each isolate

```{r}
example_data <- filter(temp_growth, ID == "example")
```

## MMRT 1.5

The below chunks fit the modified Macromolecular Rate Theory model on one example isolate. Optimum growth temperature (topt) and temperature inflection point (Tinf) are estimated by calculating the first and second derivatives, respectively,

### EXAMPLE ISOLATE

Fitting modified MMRT

```{r}
# mutating temperature to be in Kelvin if needed

example_data <- example_data %>% mutate("Temp_K" = Temp + 273.15)
rownames(example_data) <- 1:nrow(example_data)
```

```{r}
x <- example_data[,5]
y <- example_data[,4]

R <- 8.314
  
	fit <- nls(y ~ log(x)+23.76-(h/(R*x))-((((a*(x-296.1))+b)*(x-296.1))/(R*x))+(s/R)+((((a*(x-296.1))+b)/R)*(log(x/296.1))),
		start=list(b=-3000,h=10000,s=-150,a=0),control=nls.control(maxiter=100))
	# extract the parameters from the fitted model
	fit.b=coefficients(fit)[[1]]
	fit.h=coefficients(fit)[[2]]
	fit.s=coefficients(fit)[[3]]
	fit.a=coefficients(fit)[[4]]
	# create a synthetic temperature vector that we use to predict rates with the fitted model
	xs<-seq(0+273.15, 60+273.15, length.out=10000);
	# generate rate predictions using this synthetic temperature vector (we also use this to draw the curve in the plot)
	fit.pred=log(xs)+23.76-(fit.h/(R*xs))-((((fit.a*(xs-296.1))+fit.b)*(xs-296.1))/(R*xs))+
			(fit.s/R)+((((fit.a*(xs-296.1))+fit.b)/R)*(log(xs/296.1)))
```

Calculating topt and tinf
```{r}
## First derivative for calculation for Topt ##

d <- function(x) {1/x + coef(fit)[2] * R/(R * x)^2 - ((coef(fit)[4] * (x - 296.1) + ((coef(fit)[4] * (x - 296.1)) + coef(fit)[1]))/(R * 
                                                                                                            x) - (((coef(fit)[4] * (x - 296.1)) + coef(fit)[1]) * (x - 296.1)) * R/(R * x)^2) + (coef(fit)[4]/R * (log(x/296.1)) + (((coef(fit)[4] * (x - 296.1)) + coef(fit)[1])/R) * (1/296.1/(x/296.1)))}
    
# plotting the first derivative to identify max/min values
    
curve(d, from = min(x), to = max(x)); abline(h = 0, lty = 3)
    
# extracting the exact max value
    
uniroot(d, lower = 302, upper = 310) # lower and upper bounds are determined through the plots of the derivative
    
#TO <- uniroot(d, c(min(x), max(x)))$root

  
## Second derivative calculation for Tinf ##

d <- function(x) {coef(fit)[4]/R * (1/296.1/(x/296.1)) + (coef(fit)[4]/R * (1/296.1/(x/296.1)) - (((coef(fit)[4] * (x - 296.1)) + coef(fit)[1])/R) *

      (1/296.1 * (1/296.1)/(x/296.1)^2)) - (coef(fit)[2] * R * (2 * (R * (R * x)))/((R * x)^2)^2 + 1/x^2 + ((coef(fit)[4] + coef(fit)[4])/(R * x) -

        (coef(fit)[4] * (x - 296.1) + ((coef(fit)[4] * (x - 296.1)) + coef(fit)[1])) * R/(R * x)^2 - ((coef(fit)[4] * (x - 296.1) + ((coef(fit)[4] * (x - 296.1)) + coef(fit)[1])) * R/(R * x)^2 -

          (((coef(fit)[4] * (x - 296.1)) + coef(fit)[1]) * (x - 296.1)) * R * (2 * (R * (R * x)))/((R * x)^2)^2)))}

# plotting the second derivative to identify max/min values

curve(d, from = min(x), to = max(x)); abline(h = 0, lty = 3)
    
TI <- uniroot(d, c(min(x), max(x)), extendInt ="yes")$root
```

Plotting the fitted modified MMRT with the raw data of log transformed growth rate over temperature
```{r}
# creating a data frame with the predicted values
pred <- data.frame(xs, fit.pred)
# converting predicted temperature data to ºC
pred <- pred %>% mutate(Temp = xs - 273.15)
# subsetting data frame
pred <- pred %>% filter(Temp <= max(example_data$Temp) &
                          Temp >= min(example_data$Temp))

# plot
ggplot() +
  geom_point(aes(Temp, ln_r), data = example_data, alpha = 0.7) +
  geom_line(aes(Temp, fit.pred), data = pred, col = 'blue', size = 0.5) +
  theme_bw() +
  xlab("Temperature (ºC)") +
  ylab("ln(growth rate)") +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 15))
```
